---
title: 'Appendix B: growth inhibition'
author: "Jennifer Tran"
date: "2024-07-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load required packages and import data

```{r}
require("pacman")

p_load(data.table, dplyr, tidyr, purrr, DescTools)

```


I typically pre-format data from the Excel output of our Tecan MPlex.
Here, I've appended about 26 plates worth of growth curves together for ease
of importing into R. You can view a glimpse of the data here

```{r echo=FALSE}
setwd("C:/Users/jentr/OneDrive - UW-Madison/lab things/data/growth curves")

df <- fread("antibiotic_inhibition_data.csv", header=TRUE)

```

```{r, echo=FALSE}
glimpse(head(df[,1:10], n=10))

```

...but it requires some extra data wrangling as a result 

```{r}
split_df_by_plate <- function(df) {
  plate_numbers <- unique(gsub("(_.*)|(_time)", "", names(df)))
  plate_dfs <- map(plate_numbers, function(plate) {
    df %>%
      select(matches(paste0(plate, "_"))) %>%
      rename_with(~ gsub(paste0(plate, "_"), "", .), -matches(paste0(plate, "_time"))) %>%
      rename(time = matches(paste0(plate, "_time")))
  })
  names(plate_dfs) <- plate_numbers
  plate_dfs
}

```

## Function for subtracting off background

Subtract off background absorbance values from media only
Media only controls here also include the antibiotic concentration as some
antibiotics actually have additional noise at OD600/A600.
In two cases, media only controls got contaminated towards the end of the 
growth curve. In those cases, controls from the antibiotic concentration one 
higher or lower were used.


```{r}
subtract_blanks <- function(plate_df) {
  blank_cols <- grep("blank", names(plate_df), value = TRUE)
  plate_df <- plate_df %>%
    mutate(across(-time, ~ . - rowMeans(across(all_of(blank_cols)))))
  plate_df <- plate_df %>%
    select(-matches("blank"))
  plate_df
}
```

## Function for calculating area under the curve

This uses the base AUC() function from the pracma package.
It defaults to the trapezoidal method.

```{r}
calculate_auc <- function(plate_df) {
  plate_df %>%
    pivot_longer(-time, names_to = c("antibiotic", "concentration", "replicate"), names_sep = "_") %>%
    group_by(antibiotic, concentration, replicate) %>%
    summarise(auc = AUC(time, value), .groups = 'drop')
}
```

## Function for calculating percent inhibition

Each plate contains a replicate control with no antibiotic.
Nearly every plate actually use a replicate control in the same column as the
corresponding antibiotic sample.


Take the average for each antibiotic concentration across the three replicates.
The standard deviation can give an idea of how much the growth curves varied
between replicates.

```{r}
calculate_percent_inhibition <- function(auc_df) {
  control_auc <- auc_df %>% filter(concentration == 0) %>% select(antibiotic, replicate, auc)
  auc_df <- auc_df %>%
    left_join(control_auc, by = c("antibiotic", "replicate"), suffix = c("", "_control")) %>%
    mutate(percent_inhibition = 100 * (1 - (auc / auc_control))) %>%
    select(-auc_control)
  auc_df
}

summarize_percent_inhibition <- function(inhibition_df, plate_number) {
  inhibition_df %>%
    group_by(antibiotic, concentration) %>%
    summarise(
      mean_percent_inhibition = mean(percent_inhibition),
      sd_percent_inhibition = sd(percent_inhibition),
      .groups = 'drop'
    ) %>%
    mutate(plate_number = plate_number)
}
```

## Process
```{r}
process_growth_curves <- function(df) {
  plate_dfs <- split_df_by_plate(df)
  inhibition_summaries <- map2_dfr(plate_dfs, names(plate_dfs), function(plate_df, plate_number) {
    plate_df <- subtract_blanks(plate_df)
    auc_df <- calculate_auc(plate_df)
    inhibition_df <- calculate_percent_inhibition(auc_df)
    summarize_percent_inhibition(inhibition_df, plate_number)
  })
  return(inhibition_summaries)
}

result_df <- process_growth_curves(df)

```

```{r echo=FALSE}
glimpse(result_df)
```

I like to clean it up by removing the no antibiotic controls.
They're all zero percent inhibition because of how inhibition was calculated.

```{r}
cleaned_results <- result_df %>% filter(concentration != 0)
```